---
title: Debugging Tools
subtitle: "Bio302 - Biological data analysis II"
author: 
- name: "Richard J. Telford"
institute: Department of Biology, University of Bergen
date: '2016-05-05'
output: 
 ioslides_presentation:
  widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Debugging

Detective work

Imagination needed

Form a hypothesis about what is going wrong - experiment

Tools available to help identify the problem

## Syntax errors

Code not accepted by R

Brackets & commas often the problem

Read the error message - helps to locate the problem


```{r, eval = FALSE}
with(bird1 plot(weight, temp, pch = 2, col = 2)
```
```{r, eval = FALSE}
ggplot(bird1, aes(temp, weight))
+ geom_point()
```

```{r, eval=FALSE}
sapply(bird1, function(x){mean(x))
```


## Debugging tools

`print`

`browser`

`debug` / `undebug`

`traceback`

`options(error = recover)`

## `traceback()`
Prints the stack of functions called. 

Helps identify where the error first occurred


## `print`

Look at the objects (or some aspect of them)

Very useful

## `browser`

Inserting `browser()` into a function starts the debugger

## `debug`
If you don't want to or cannot put `browser` into a function use

`debug(foo)` where `foo` is the function you want to debug

Opens the debugger on `foo` next time it is called

Use `undebug(foo)` to turn it off

## `options(error = recover)`

Chance to look at the state of all functions when the error occurred

Very useful

Turn off with `options(error = NULL)`

## Warnings

Not necessarily benign

Can be difficult to track down

By default not printed until calculations completed

Promote to errors with

`options(warn = 2)`

## Getting help

`?lm` opens help file on function `lm`

`example(lm)` runs the examples from `lm`

`??regression` finds help about regression

R-coder club - Fridays at 1400

Stackoverflow.com

 - Check for existing answers
 - Minimal example
 - Make problem reproducible
 - Minimal dataset
 `dput` or make random data (set seed first)