---
title: Generalised Least Squares<br>&<br>Non-linear Least Squares
subtitle: "Bio302 - Biological data analysis II"
author: 
- name: "Richard J. Telford"
institute: Department of Biology, University of Bergen
date: '2016-05-05'
output: 
 ioslides_presentation:
  widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(nlme)
```

## Assumptions of Least Squares

1. The relationship between the response and the predictors is ~linear.
2. The residuals have a mean of zero.
3. __The residuals have constant variance (not heteroscedastic).__
4. The residuals are independent (uncorrelated).
5. The residuals are normally distributed.

Weighted least squares can be used when variance is not constant


## Weighted least squares

If observations have different amounts of uncertainty

Give uncertain observation less weight 

 – less influence on regression

Adjust weight matrix

##
W = weight matrix

k = sqrt(W)

```{r, echo = FALSE}
diag(4)
```

Default weight matrix
 - 1s on diagonal
 - 0s off diagonal

y[i] = β[0] + β[1]x[i] + ε[i]

k[i]y[i] = k[i]β[0] + k[i]β[1]x[i] + k[i]ε[i]

We adjust the influence of the point

'We adjust the size of the residuals'

## Galton's Peas

Pea sizes
```{r, echo = FALSE}
peas <- read.table(header = TRUE, text = "
Parent	Progeny	SD
0.21	0.1726	0.01988
0.2	0.1707	0.01938
0.19	0.1637	0.01896
0.18	0.164	0.02037
0.17	0.1613	0.01654
0.16	0.1617	0.01594
0.15	0.1598	0.01763
")
knitr::kable(peas)
```

##

 - Parent - pea diameter in inches of parent plant 
 - Progeny - mean pea diameter (inches) of up to 10 plants grown from seeds of the parent plant
 - SD - standard deviation of offspring peas grown from each parent. 
 
Standard deviations reflect information in the response 

Downweight obervations with large standard deviation and upweight observations with a small standard deviation. 

Use weighted least squares with weights equal to 1/SD^2^. 

## In R

```{r}
mod <- lm(Progeny ~ Parent, data = peas)
summary(mod)$coef
wmod <- lm(Progeny ~ Parent, data = peas, weights = 1 / SD ^ 2)
summary(wmod)$coef
```

Weighed regression assumes that the weights are precisely known

##
```{r}
ggplot(peas, aes(x = Parent, y = Progeny, size = 1 / SD ^ 2)) + 
  geom_point() +
  geom_abline(intercept = coef(mod)[1], slope = coef(mod)[2], col = "black") +
  geom_abline(intercept = coef(wmod)[1], slope = coef(wmod)[2], col = "red") +
  labs(size = "Weight")
```

## Generalised least squares

Solution to non-constant variance (heteroscedastic data)

W = 1/var(response)

Fit using `gls` in package `nlme`.

## Bird1 dataset{.columns-2}
```{r, echo = FALSE}
bird1 <- read.csv("data/bird1.csv")
```
Variances are different for each treatment
```{r, fig.height=3.5, fig.width=5}
ggplot(bird1, aes(x = treat, y = weight)) +
  geom_boxplot()
with(bird1, tapply(weight, treat, var))
```

## GLS{.smaller}

```{r}
library(nlme)
fit_gls <- gls(weight ~ treat, data = bird1, weights = varIdent(form = ~ +1|treat))
summary(fit_gls)
```

## Model comparison
```{r}
fit_0 <- gls(weight ~ treat, data = bird1)
anova(fit_0, fit_gls)
```

## Other magic in GLS{.smaller}

Copes with spatially/temporally autocorrelated data 

  
```{r, fig.height=5, fig.width=7}
year <- time(LakeHuron)
par(mfrow = c(1, 2))
acf(resid(lm(LakeHuron ~ year)))
pacf(resid(lm(LakeHuron ~ year)))
par(mfrow = c(1, 1))
```
##
Many different correlation structures available. Choose appropriate one.

Change off-diagonal values in weights-matrix
```{r}
mod_ar1 <- gls(LakeHuron ~ year, corr=corAR1())
summary(mod_ar1)
```

# Non-linear least squares

## Assumptions of Least Squares

1. __The relationship between the response and the predictors is ~linear.__
2. The residuals have a mean of zero.
3. The residuals have constant variance (not heteroscedastic).
4. The residuals are independent (uncorrelated).
5. The residuals are normally distributed.

Nonlinear least squares can be used when relationship is non-linear.

## Non-linear least squares{.columns-2}

We know something about the relationship

Upper or lower bound (asymptote)

```{r, echo=FALSE, fig.width=5, fig.height = 3.5}
amount <- read.csv("data/amount.csv")
ggplot(amount, aes(x = calcium, y = amount)) +
  geom_point()
```

Three options

 - Transform response
 - Use polynomials (x2)
 - Non-linear expression

## 
Transformations:

- AVOID, statistics must be correct on transformed scale
- Log transform assumes that error is a constant proportion of the response

##
Polynomial

y[i] = β[0] + β[1]x[i] + β[2]x[i]^2^ + ε[i]

 - Very useful
 - But can give incorrect predictions
 
 ```{r, echo=FALSE, fig.width=5, fig.height = 3.5}
ggplot(amount, aes(x = calcium, y = amount)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x + I(x^2))
```

## NLS
The relationship has an asymptote, and a exponential decline

yi = β0 + β1 exp(β2xi) + εi

 - β0  = asymptote
 - β0 + β1 = intercept
 - β2 = slope (proportional change)

If x -> inf & β2<0 
- 		β1 exp(β2xi) = 0

If x -> 0
 - 		β1 exp(β2xi) = β1

##
```{r}
library(nlme)
fit.nls<-nls(amount ~ b0 + b1 * exp(b2 * calcium), data = amount, 
             start=c(b0=0,b1=20,b2=-1))
summary(fit.nls)
```

##
```{r}
ggplot(cbind(amount, fit = fitted(fit.nls)), aes(x = calcium, y = amount)) +
  geom_point() +
  geom_line(aes(y = fit))
```

##
1. Is there a natural expression
2. Identify parameters
3. Roughly estimate parameters
4. Start NLS