---
title: Lubridate
subtitle: "Bio302 - Biological data analysis II"
author: 
- name: "Richard J. Telford"
institute: Department of Biology, University of Bergen
date: '2016-05-05'
output: 
 ioslides_presentation:
  widescreen: true
---

```{r setup_im, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE)
```

## Importing data

Critical step in any analysis

Often difficult

For delimited files (eg. csv)

  try `readr` package
  
For excel files

  try `readxl` package

##readr

Replacements for base functions

- `read.table` -> `read_table`
- `read.csv` -> `read_csv`

Advantages over base functions
 
 - can read compressed files
 - faster
 - does not convert strings to factors (except on request)
 - column types guessed or set by user
 - gives table of problems

[https://blog.rstudio.org/2015/04/09/readr-0-1-0/](https://blog.rstudio.org/2015/04/09/readr-0-1-0/)

##`read_excel` in `readxl`

Many people keep data in excel format (probably unwise)

Nicely formatted data easy to read with `read_excel`

```{r, eval = FALSE}
library(readxl)
df <- read_excel(path = "myExcel", sheet = "Sheet1")
```

Returns a tibble (tbl_df), convert to a pure `data.frame` with `as.data.frame`.

## Exercise

Use `readxl` package to import an excel file 