---
title: "Makerere R workshop"
author: 
- name: "Richard J. Telford"
institute: Department of Biology, University of Bergen
date: "October 26, 2016"
output: 
 ioslides_presentation:
  widescreen: true
---

```{r setup0, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4, fig.width = 5, cache = TRUE, message = FALSE, cache.lazy = TRUE, eval = TRUE)
```

## Rationale

 - Introduce Makerere staff & students to recent developments in R
 - Focus on staff & students with existing R skills who train others
 
All code and presentations at [https://github.com/richardjtelford/Bio302](https://github.com/richardjtelford/Bio302)

 
## Contents

 - Integrated Development Environments - Rstudio
 - Style guide for readable code
 - `readxl` and `readr` for importing data
 - `ggplot2` for better graphics including maps with `ggmaps`
 - `dplyr` and `tidyr` for data manipulation
 - `lubridate` for date manipulation
 - Debugging R code
 - version control - use git (and github)
 - Rmarkdown for reproducible analyses
 - Online resources at the University of Bergen

# Use an Integrated Development Environment

```{r child = "useAnIDE.Rmd"}
``` 

# You need a style guide

```{r child = "styleguide.Rmd"}
``` 

# Importing data

```{r child = "importing.Rmd"}
``` 


# ggplot2 and allied packages

```{r child = "ggplot_etc.Rmd"}
``` 

# lubridate for handling dates

```{r child = "lubridate.Rmd"}
``` 

# Debugging R

```{r child = "DebuggingTools.Rmd"}
``` 

# Rmarkdown

```{r child = "rmarkdown.rmd"}
``` 
