---
title: "RMarkdown"
author: "Richard J. Telford"
date: "May 25, 2016"
output: ioslides_presentation
---

```{r setup_rmd, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Why

 - Interlaced R code and text compiles to make presentation/paper/thesis 
 - Dynamic documents - update when data/code changes
 - Reproducible analyses
 - Simple formatting
 - Text files - good for version control

## Output formats

Documents

 - PDF
 - HTML
 - word

Presentations

 - ioslides

## Formatting

Standard markdown formatting

\*italic\* or \_italics\_ -> *italic*

\*\* bold\*\* or \_\_bold\_\_ -> **bold** 

\#Title

\#\#Subtitle (new page in ioslides)

\#\#\#Sub-subtitle 

\[Link text](biostats.b.uib.no) -> [Link text](biostats.b.uib.no)

\!\[Image description](image.jpg)  -> includes image.jpg
 
Rmarkdown cheat sheet
[https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf
](https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf)


## R chunks

R code in chunks

\```{r chunkName, eval = TRUE}  
#some R code  
library(vegan)  
x <- 2  
y <- 7  
\```

Chunk name is optional.

Chunk options

 - `eval` Run the R code
 - `echo` Show the R code

Full list [http://yihui.name/knitr/options/](http://yihui.name/knitr/options/)

## Inline code

Seven times six is `` `r '\x60r 7 * 6\x60'` ``

Seven times six is `r 7 * 6`

## Nice tables

```{r, echo = TRUE}
knitr::kable(head(iris))
```

## Getting stated

In RStudio

`File>>New File>>R Markdown...`

Select options for document format

R gives example document to edit

Click `Knit` to render

## Advanced

Can include bibliographies - [rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html](http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html)

Latex/html/css to change formatting

Cache options to avoid recalculating expensive objects

Use `knitr::read_chunk` to read code from another file.

Can include citations and automatically formatted reference list. 

## Reading list

Reproducible Science with Rmarkdown
http://pakillo.github.io/Rmarkdown_talk_SevillaR_Nov2014/#/

Rmarkdown reference guide https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf

Rmarkdown cheat sheet
https://www.rstudio.com/wp-content/uploads/2016/03/rmarkdown-cheatsheet-2.0.pdf

