---
title: Adopt a Style Guide
subtitle: "Bio302 - Biological data analysis II"
author: 
- name: "Richard J. Telford"
institute: Department of Biology, University of Bergen
date: '2016-05-05'
output: 
 ioslides_presentation:
  widescreen: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Why

 - Makes code easier to read
 - Makes code easier to debug

Make your own style - but be consistent

## Naming Things

>    “There are only two hard things in Computer Science: cache invalidation and naming things.”

>    — Phil Karlton

## Naming Objects

 - Names can contain letter numbers "_"" and "."
 - Names must begin with a letter or "."
 - Avoid using names of existing functions - confusing
 - Make names concise yet meaningful
 - Reserved words include TRUE, for, if
 
## Which of these are valid names?

 - min_height
 - max.height 
 - _age 
 - .mass 
 - MaxLength 
 - min-length 
 - FALSE
 - 2widths 
 - celsius2kelvin 
 - plot

 
## Spacing

White-space is free!

Place spaces around all infix operators (=, +, -, <-, etc.) and around = in function calls. 

Always put a space after a comma, and never before.

### Good
```{r}
average <- mean(feet / 12 + inches, na.rm = TRUE)
ChickWeight[1, ]
```
### Bad
```{r}
average<-mean(feet/12+inches,na.rm=TRUE)
ChickWeight[1,]
```
##
Exception: :, :: and ::: don’t need spaces around them.

### Good
```{r}
x <- 1:10
base::get
```

### Bad

```{r}
x <- 1 : 10
base :: get
```
:: notation tells R which package to use


## Split long commands over multiple lines

```{r}
list(
  total = a + b + c, 
  mean  = (a + b + c) / n
)
```

## Indentation makes code more readable

### Good
```{r}
pos <- function(x) {
  if (is.null(dim(x))) {
    x[x > 0]
  } else{
    x[, colSums(x) > 0, drop  = FALSE]
  }
}
```

### Bad
```{r}
pos <- function(x){
if(is.null(dim(x)))
{x[x >0]} 
else{
x[, colSums(x) > 0, drop  = FALSE]
}}
```

## Assignment

Use <-, not =, for assignment.

### Good
```{r}
x <- 5
```
### Bad
```{r}
x = 5
```


## Comments

Use # to start comments.

Help you and others to understand what you did

Comments should explain the why, not the what.

Use comments to break up file into readable chunks for navigation
```{r}
#### Load data ####

#### Plot data ####
```
 
## Split analyses over multiple files

Long scripts become difficult to navigate

Fix by moving parts of the code into different files

For example: 
 - data import code to "loadData.R"
 - functions to "functions.R"

Import with  
```{r}
source("loadData.R")
source("functions.R")
```
 
## Repeated code

Make repeated code into functions.

```r{}
plot(obj1, ...)
legend(...)

plot(obj2, ...)
...

plotObj <- function(obj){
  plot(obj, ...)
  legend(...)
}

plotObj(obj1)
```
Single place to maintain
 
## Avoid `attach()`

Unless you like strange bugs

Rarely useful to attach - many better options

https://coderclub.b.uib.no/2016/05/03/dont-get-attached-to-attach/
 
## Reading

Google's R Style Guide [https://google.github.io/styleguide/Rguide.xml](https://google.github.io/styleguide/Rguide.xml)

Hadley Wickham, H. Style Guide _Advanced R_
[http://adv-r.had.co.nz/Style.html](http://adv-r.had.co.nz/Style.html)

## Exercise

Take some of your existing R code and reformat it according to these style guides. 

Check your code still works!