---
title: Use an Integrated Development Environment
subtitle: "Bio302 - Biological data analysis II"
author: 
- name: "Richard J. Telford"
institute: Department of Biology, University of Bergen
date: '2016-05-05'
output: 
 ioslides_presentation:
  widescreen: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Why

A good IDE makes R coding more efficient and reduces some types of bugs

 - Bracket matching
 - Autocomplete of function/variable names
 - Integrated help
 - Project management and version control
 - Help debugging

## Choices

 - Tinn-R (Windows only)
 - RKWard
 - emacs
 - __RStudio__ [rstudio.com](https://www.rstudio.com)

## Demonstration

 - Autocompletion
 - Bracket matching
 - Code formatting
 - alt-shift-k to get shortcuts
 
## Project management
  
Use Rstudio projects for project management

 - Keeps related files together
 - Keeps track of working directory, History and environment
 
## Exercise

Set up an Rstudio project called Bio302

We will use this project all week.

Open a new script and start to __type__ some code

```{r}
set.seed(42) # make reproducible
df <- data.frame(
  x = 1:10,
  y = rnorm(10)
)

plot(df$x, df$y)
legend("topright", legend = "Random", pch = 1)

```

Become familiar with the autocomplete, sending code to the console and some shortcuts
