---
title: "Untitled"
author: "Richard J. Telford"
date: "June 8, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Exercise Diagnostics

1. Import sorich.csv

2. Fit a linear model for nsp as a function of cover

3. Examine the diagnostics graphs. What suggests this may not be an appropriate model to fit?

4. Import bird1.csv

```{r}
library("tidyverse")
bird <- read_csv("data/bird1.csv")
```


5. Fit a linear model for weight as a function of temp

```{r}
mod1 <- lm(weight ~ temp, data = bird)
```


6. Examine the diagnostics graphs. Do they indicate that any action needs to be taken.

```{r}
library("ggfortify")
autoplot(mod1)
```


7. Fit a model that includes a quadratic term for temp 

```{r}
mod2 <- lm(weight ~ temp + I(temp^2), data = bird)
```


8. Compare the diagnostics plots with those from the previous model. Comment on any changes.

```{r}
autoplot(mod2)
```


9. Extract the AIC from both  models (hint use function AIC). Which is the better model?

```{r}
AIC(mod1, mod2)
```


10. Calculate the deltaAIC for each model

11. Calculate the AIC weights for each model. Interpret these weights.

```{r}
 AICcmodavg::aictab(list(mod1, mod2))
```



12. With the built-in data set `swiss` (use `data(swiss)`), make ols model between Fertility and all other variables. Find the VIF of each predictor. Are there any problem variables?
```{r}
mod3 <- lm(Fertility ~ ., data = swiss)
olsrr::ols_vif_tol(mod3)
swiss %>% select(-Fertility) %>% pairs(gap = 0)
```

13. Use `MASS::mvrnorm()` to simulate 100 observation of two predictor variables (x and z) with a given correlation. Simulate a response variable y = b0 + b1x + b2z. Test how the uncertainty in the coefficients changes with the correlation (and hence vif) of the predictor variables.



```{r}
library("broom")

#dplyr function
analyse <- . %>%   #can reuse code
  map(as.data.frame) %>% 
  map(rename, x = V1, z = V2) %>% 
  map(mutate, y = x + z + rnorm(n)) %>% #true effect size is 1 for both predictors
  map(~lm(y ~ x + z, data = .)) %>% 
  map_df(tidy) %>% 
  filter(term != "(Intercept)")

n <- 100
nrep = 100
cor_matrix0 <- matrix(c(1, 0, 0, 1), nrow = 2)
cor_matrix9 <- matrix(c(1, 0.9, 0.9, 1), nrow = 2)
cor_matrix99 <- matrix(c(1, 0.99, 0.99, 1), nrow = 2)

bind_rows(
 r0 = rerun(.n = nrep, {
    MASS::mvrnorm(n = n, mu = rep(0, 2), Sigma = cor_matrix0)
  }) %>%
  analyse(),

  r9 = rerun(.n = nrep, {
    MASS::mvrnorm(n = n, mu = rep(0, 2), Sigma = cor_matrix9)
  }) %>%
  analyse(),

  r99 = rerun(.n = nrep, {
    MASS::mvrnorm(n = n, mu = rep(0, 2), Sigma = cor_matrix99)
  }) %>%
  analyse(), 
 .id = "correlation"
) %>% 
  ggplot(aes(x = correlation, y = estimate, fill = correlation)) + 
  geom_violin()


```

