---
title: "Spatial and temporal structure"
author: "Richard J. Telford"
date: "June 1, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE)
```

1) Simulate a 100-observation autocorrelated timeseries with `arima.sim`, with a first order autoregressive coefficient of 0.5. Also make a time vector of `1:100`

```{r}
library("tidyverse")
y <- arima.sim(model = list(ar = 0.5), n = 100)
plot(y)#base plot
#or for a ggplot, use autoplot (many autoplot functions in ggfortify package)
library("ggfortify")
autoplot(y)#uses fortify to convert ts object to a data.frame 


set.seed(42)
df <- tibble(
  x = 1:100,
  y = arima.sim(model = list(ar = 0.5), n = 100)
)
```

2) Plot the data. 

```{r}
ggplot(df, aes(x = x, y = y)) + geom_line()
```



3) Regress the timeseries against time with an OLS model. Does the model appear to be statistically significant?

```{r}
mod <- lm(y ~ x, data = df)
anova(mod)
```


4) Plot the model diagnostics, including an `acf` and `pacf` of the residuals.

```{r}
library("ggfortify")
autoplot(mod)

acf(resid(mod))#use autoplot to get ggplot version
pacf(resid(mod)) # single spike
```



4) Use the Durbin-Watson test to test the residuals for autocorrelation.

```{r}
library("lmtest")
dwtest(mod)

```


5) Fit a gls with an appropriate correlation structure. Is this a better model? How have the p-value and effect size changed?

```{r}
library("nlme")

mod2 <- gls(y ~ x, data = df, correlation = corAR1())
anova(mod2)
list(mod, mod2) %>% map(coef) %>% map("x") # should be unbiased

```



5) Repeat the above 1000 times and find how autocorrelation affects the distribution of the p-value and the effect size.


```{r}
library("broom")
library("broom.mixed") # expansion package for broom 
#no autocorrelation

set.seed(1)
sim_lm0 <- rerun(.n = 1000, tibble(
  x = 1:100,
  y = rnorm(100)
)) %>% 
  map(~lm(y ~ x, data = .)) %>% 
  map_df(tidy) %>% 
  filter(term == "x") %>% 
  mutate(sig = p.value < 0.05)


set.seed(1)
sim_lm <- rerun(.n = 1000, tibble(
  x = 1:100,
  y = arima.sim(model = list(ar = 0.5), n = 100)
)) %>% 
  map(~lm(y ~ x, data = .)) %>% 
  map_df(tidy) %>% 
  filter(term == "x") %>% 
  mutate(sig = p.value < 0.05)


set.seed(1)
sim_gls <- rerun(.n = 1000, tibble(
  x = 1:100,
  y = arima.sim(model = list(ar = 0.5), n = 100)
)) %>% 
  map(~gls(y ~ x, data = ., correlation = corAR1())) %>% 
  map_df(tidy) %>% 
  filter(term == "x") %>% 
  mutate(sig = p.value < 0.05)


all_models <- bind_rows(list(lm_ar0 = sim_lm0,
               lm_ar0.5 = sim_lm,
               gls_ar0.5 = sim_gls), .id = "model") %>% 
  mutate(model = factor(model, levels = c("lm_ar0", "lm_ar0.5", "gls_ar0.5")))


nbins <-  20
ggplot(all_models, aes(x = p.value)) +
  geom_histogram(binwidth = 1/nbins, center = 1/nbins/2) +
  geom_hline(yintercept = 1000/nbins, colour = "red") +
  facet_wrap(~ model, ncol = 1)


ggplot(all_models, aes(x = estimate, fill = sig)) +
  geom_histogram() +
  geom_vline(xintercept = 0, colour = "grey50", linetype = "dashed") +
   facet_wrap(~ model, ncol = 1)
```




## Real data

1) The built-in dataset LakeHuron has annual lake level data from 1885 to 1972
Load the data with the command data(LakeHuron)

```{r}
data("LakeHuron")
Huron <- tibble(year = time(LakeHuron), level = as.vector(LakeHuron))
```

2) Plot the data.

```{r}
ggplot(Huron, aes(x = year, y = level)) + 
  geom_line()
```


3) Regress the LakeHuron lake level against year using a linear model. Is there a significant trend?

```{r}
mod <- lm(level ~ year, data = Huron)
summary(mod)

```



4) Plot the autocorrelation and partial autocorrelation functions for the residuals from this regression. Interpret them.

```{r}
acf(resid(mod))
pacf(resid(mod))
```



5) Fit an autoregressive models to the residuals. Compare the results with your interpretation of the PACF plot.

```{r}
ar(resid(mod))
```


6) Fit a gls model using a corAR1 correlation structure. Test if the correlation structure is necessary. Is the trend significant? 

```{r}
mod_gls <- gls(level ~ year, data = Huron, correlation = corAR1())
summary(mod_gls)
```



7) Fit a gls model using a corARMA correlation structure with two AR terms. Is this model an improvement?

```{r}
mod_gls2 <- gls(level ~ year, data = Huron, correlation = corARMA(p = 2))
summary(mod_gls)

anova(mod_gls, mod_gls2)
```

