---
title: "Using tidyverse with the Bergenbysykkel  data"
author: "Richard J. Telford"
date: "May 15, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Dataset 03.csv is the data on the hire of Bergenbysykkel in March 2019 from [https://bergenbysykkel.no/en/open-data/historical](https://bergenbysykkel.no/en/open-data/historical), showing the start and end times and location of each time a bicycle was hired.

## 1 Import the data

Import the data with `read_csv` and examine it.

```{r}
bike <- readr::read_csv("data/03.csv")#better import (usually) than read.csv
head(bike)
```


## 2 Explore the data
Use tidyverse and make plots when useful
You will probably need to use `group_by`, `count`, `summarise`, `mutate`, `slice` and `ggplot`.

 - Which is the most popular starting station?
 
```{r}
bike %>% group_by(start_station_name) %>% 
  count() %>% 
  ungroup() %>% 
  slice(which.max(n))

```
 
 
 - Plot the number of hires and returns from each station
 
```{r} 
bike %>% 
  select(start_station_name, end_station_name) %>% 
  gather(key = key, value = station) %>% 
  count(key, station) %>% 
  filter(n > 10) %>% #remove stations with v low usage
  mutate(key = factor(key, levels =  c("start_station_name", "end_station_name"))) %>% #force order of bars
  ggplot(aes(x = station, y = n, fill = key)) + 
  geom_col(position = position_dodge()) +
  scale_fill_discrete(labels = c("start_station_name" = "Hire", "end_station_name" = "Return")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
``` 

 - Which is the most popular pair of start and end stations?
 
```{r}
bike %>% group_by(start_station_name, end_station_name) %>% 
  count() %>% 
  ungroup() %>% 
  slice(which.max(n))

```
 
 - What was the longest/shortest duration of hire? 
```{r}
bike %>% 
  select(duration) %>% 
  slice(which.max(duration)) %>% 
  mutate(duration = duration/60/60)#hours

```

 
 - Plot the distribution of hire duration.
```{r}
bike %>% 
 ggplot(aes(x = duration/60)) +#minutes
  geom_histogram() + 
  xlim(0, 60)
```
 
 - What is the median duration of hire from each station?
```{r}
bike %>% group_by(start_station_name) %>% 
  filter(n() > 10) %>% #remove stations with very few hires
  summarise(med = median(duration)) %>% 
  arrange(med)

```

 
 - Map this information
```{r}
med_dur <- bike %>% group_by(start_station_name, start_station_latitude, start_station_longitude) %>% 
  filter(n() > 10) %>% #remove stations with very few hires
  summarise(med = median(duration)/60)  
  
ggplot(med_dur, aes(x = start_station_longitude, y = start_station_latitude, size = med)) +
  geom_point(colour = "grey50") +
  scale_size_area(name = "Median duration") +
  ggrepel::geom_text_repel(aes(label = start_station_name), size = 3)

```  
 
 - Are there any significant differences in duration between stations.
```{r}
mod_dur <- bike %>% group_by(start_station_name, start_station_latitude, start_station_longitude) %>% 
  filter(n() > 10,  #remove stations with very few hires
         duration < 60 * 60) %>% #remove outliers
  lm(duration ~ start_station_name, data = .)
anova(mod_dur)
```
 
 
 - How far does a typical cyclist travel? (assuming straight line)
 
```{r}
distances <-  bike %>%
  mutate(distance = fields::rdist.earth.vec(#long/lat so need great circle distances
    x1 = cbind(start_station_longitude, start_station_latitude),
    x2 = cbind(end_station_longitude, end_station_latitude),
    miles = FALSE
  ))
  
distances %>% 
  ggplot(aes(x = distance)) + 
  geom_histogram() + 
  labs(x = "Distance km")
 
 
 
```
 
 - What is the relationship between distance travelled and time taken?
 
```{r}
ggplot(distances, aes(x = distance, y = duration/60)) + geom_point(alpha = 0.2) + ylim(0, 60)
#could use quantile regression to find lower limit to cloud of points
```
 
 - How fast was the fastest cyclist (for simplicity assume a straight line of travel)
```{r}
distances %>% 
  mutate(speed = distance/(duration/60/60)) %>% #km/hr
  select(start_station_name, end_station_name, speed) %>% 
  arrange(desc(speed)) %>% 
  slice(1:5)#impressively fast
```
 
 
 
## Day and time


With help from the `lubridate` package
 - How does the number of hires vary throughout the day and week?
```{r}
library("lubridate")
bike %>% 
  mutate(hour = hour(started_at), day = wday(started_at, label = TRUE)) %>%
  count(day, hour) %>% 
  ggplot(aes(x = hour, y = n, colour = day)) + geom_line()
#very different weekday and weekend usage
```

 
 
 - How does the average speed vary through the day?
```{r}
distances %>% 
  filter(wday(started_at, week_start = 7) %in% 2:6) %>% #look at weekdays only
  mutate(hour = hour(started_at)) %>% 
  mutate(speed = distance/(duration/60/60)) %>%
  ggplot(aes(x = hour, y = speed, group = hour)) + geom_boxplot()
  
```
 
 
 - Are some stations more popular at different times of day for hire and return?
 
```{r}
bike %>% mutate(hour = hour(started_at)) %>%
  group_by(start_station_name, hour) %>% 
  count() %>% 
  group_by(hour) %>% 
  filter(between(hour, 5, 22)) %>% 
  mutate(percent = n/sum(n) * 100) %>% 
  group_by(start_station_name) %>% 
  filter(sum(n) > 400) %>% #remove little used stations
  ggplot(aes(x = hour, y = percent, colour = start_station_name)) + 
  geom_line()


bike %>% mutate(day = wday(started_at, label = TRUE)) %>%
  group_by(start_station_name, day) %>% 
  count() %>% 
  group_by(day) %>% 
  mutate(percent = n/sum(n) * 100) %>% 
  group_by(start_station_name) %>% 
  filter(sum(n) > 400) %>% #remove little used stations
  ggplot(aes(x = day, y = percent, colour = start_station_name, group = start_station_name)) + 
  geom_line()
```
 
## Other analyses ... 
 
Do something informative with these data. 

Packages to try `gganimate`,`ggrepel`, `ggmaps` (`get_stamenmap` will work without a google license)
 