---
title: "GLM & GAM"
author: "Richard J. Telford"
date: "May 30, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("ggplot2")
```

#Exercises GLM & GAM

##Poisson GLM
1. Import sorich.csv
```{r}
if(interactive()){
  sorich <- read.csv("data/sorich.csv")
}else{
  sorich <- read.csv("../../data/sorich.csv")
}
str(sorich)
```

2. Fit a linear model for species richness against cover assuming a Normal distribution
```{r}
mod <- lm(nsp ~ cover, data = sorich)
```

3. Study the diagnostic plots
```{r}
library(ggfortify)
autoplot(mod)
```

QQ plot look bad - heavy tail at top end
Heteroscadistic - unequal variance

4. What distribution is the response?

Discrete data - non-negative count-like. Poisson-like 

5. Do an appropriate analysis
```{r}
modP <- glm(nsp ~ cover, data = sorich, family = poisson)
```
Need to use `family = poisson` or defaults to Gaussian.

6. Check for over-dispersion
```{r}
anova(modP)
```
Residual deviance >> Residual degrees of freedom -> over dispersed

7. Interpret the results
```{r}
summary(modP)
```
Positive effect of cover on richness.

8. How does the width of the confidence interval at cover = 10 change when over dispersion is allowed for
```{r}
#Width with poisson model
predP <- predict(modP, newdata = data.frame(cover = 10), se.fit = TRUE)
predP
ciP <-with(predP, exp(fit + c(-1, 1) * 1.96 * se.fit))
ciP
diff(ciP)

#Width with poisson model
modQ <- glm(nsp ~ cover, data = sorich, family = quasipoisson)
predQ <- predict(modQ, newdata = data.frame(cover = 10), se.fit = TRUE)
ciQ <- with(predQ, exp(fit + c(-1, 1) * 1.96 * se.fit))
ciQ
diff(ciQ)

#width with negative binomial          
library("MASS") 
modN <- glm.nb(nsp ~ cover, data = sorich)
predN <- predict(modN, newdata = data.frame(cover = 10), se.fit = TRUE)
ciN <- with(predN, exp(fit + c(-1, 1) * 1.96 * se.fit))
ciN
diff(ciN)
```


9. Do the grasses differ from the herb, i.e. include the factor grasherb and test its significance?
```{r}
sorich$grasherb <- factor(sorich$grasherb)
mod2 <- glm.nb(nsp ~ cover + grasherb, data = sorich)
anova(modN, mod2)
AIC(modN, mod2)
mod3 <- glm.nb(nsp ~ cover * grasherb, data = sorich)
anova(mod2, mod3)
AIC(mod2, mod3)

preds <- predict(mod3, se.fit = TRUE)

sorich2 <- with(preds, cbind(
  sorich,
  fit = exp(fit),
  up = exp(fit + 1.96 * se.fit),
  down = exp(fit - 1.96 * se.fit)
  ))

ggplot(sorich2, aes(x = cover, y = nsp, colour = grasherb)) +
  geom_jitter(height = 0.4, width = 0.3, alpha  = 0.5) +
  geom_line(aes(y = fit)) +
  geom_ribbon(aes(ymax = up, ymin = down), alpha = 0.2, linetype = 0)
```



## How much does over-dispersion affect results
1. Use `sample()` to randomise the response in sorich
```{r}
sorich_mixed <- sorich
sorich_mixed$cover <- sample(sorich_mixed$cover)
```

2. Test if a Poisson GLM with cover as a predictor is significant
```{r}
modx <- glm(nsp ~ cover, data = sorich_mixed, family = poisson)
anova(modx, test = "Chi")
anova(modx, test = "Chi")$Pr[2]
```

3. Repeat 1000 times with `replicate`
```{r}
res <- replicate(1000, {
  sorich_mixed$cover <- sample(sorich_mixed$cover)
  modx <- glm(nsp ~ cover, data = sorich_mixed, family = poisson)
  anova(modx, test = "Chi")$Pr[2]
})
ggplot(data.frame(res), aes(x = res)) + 
  geom_histogram(binwidth = 0.05, center = 0.025) + 
  geom_hline(yintercept = 50)
sum(res < 0.05)

#repeat for quasibinomial
resQ <- replicate(1000, {
  sorich_mixed$cover <- sample(sorich_mixed$cover)
  modx <- glm(nsp ~ cover, data = sorich_mixed, family = quasipoisson)
  anova(modx, test = "F")$Pr[2]
})
ggplot(data.frame(resQ), aes(x = resQ)) + 
  geom_histogram(binwidth = 0.05, center = 0.025) + 
  geom_hline(yintercept = 50)
sum(resQ < 0.05)
```

## Binomial GAM

1. Open library mgcv
```{r}
library("mgcv")
```

2. Import data pot.csv
```{r}
if(interactive()){
  path <- ""
}else{
  path <- "../../"
}
pot <- read.csv(paste0(path, "data/pot.csv"))
summary(pot)
```

3. What type of distribution is the response variable?
0 or 1 only. Success-failure. Binomial

4. What type of link-function do we use?
logit

5. Do an appropriate GLM analysis?
```{r}
ggplot(pot, aes(x = alt, y = potalp)) +
  geom_point()

mod <- glm(potalp ~ alt, data = pot, family = binomial)
anova(mod, test = "Chi")
summary(mod)$coef

mod2 <- glm(potalp ~ alt + I(alt^2), data = pot, family = binomial)
anova(mod, mod2, test = "Chi")
summary(mod2)
```

6. Interpret the results?
Second order polynomial model significant. First order not. alt^2 coeficient negative - Unimodal relationship

7. Do a GAM analysis
```{r}
mod_gam <- gam(potalp ~ s(alt), data = pot, family = binomial)
```

8. Compare the GLM and GAM models.
```{r}
mod_glm <- gam(potalp ~ alt + I(alt^2), data = pot, family = binomial)
anova(mod_glm, mod_gam, test = "Chi")
AIC(mod_glm, mod_gam)

```

9. Which model would you prefer, and why?
glm. Simpler, not sigificantly worse, lower AIC. Easier to interpret.

```{r}
ggplot(cbind(pot, gam = fitted(mod_gam), glm = fitted(mod_glm)), aes(x = alt, y = potalp)) +
  geom_point() +
  geom_line(aes(y = gam, colour = "GAM")) +
  geom_line(aes(y = glm, colour = "GLM")) +
  labs(x = "Elevation", y = "P. alpina", colour = "Model")
```

