---
title: "Mixed Effect Models"
author: "Richard J. Telford"
date: "June 1, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## R Markdown
Exercises 

1) Import data LME.csv <br>
id gives the group id, y is the response, ab and AB are predictors

```{r}
library("ggplot2")
library("gridExtra")
if(interactive()){
  path <- ""
}else{
  path <- "../../"
}
LME <- read.csv(paste0(path, "data/LME.csv"))
summary(LME)

LME$ab <- as.factor(LME$ab)
LME$AB <- as.factor(LME$AB)

g_ab <- ggplot(LME, aes(x = ab, y = y)) + geom_boxplot()
g_AB <- ggplot(LME, aes(x = AB, y = y)) + geom_boxplot()
grid.arrange(g_ab, g_AB, nrow = 1)
```

2) What is a random effect?

3) What is a fixed effect?

4) Fit an appropriate a model with y and AB
```{r}
library("nlme")
mod1 <- lme(y ~ AB, data = LME, random =  ~ 1 | id)

library("lme4")
mod1b <- lmer(y ~ AB + (1|id), data = LME)

```


5) Interpret the results

```{r}
anova(mod1)
summary(mod1)

anova(mod1b)
summary(mod1b)
```


6) Fit an appropriate a model with y and ab
```{r}
mod2 <- lme(y ~ ab, data = LME, random =  ~ 1 | id)

```

7) Interpret the results
```{r}
anova(mod2)
summary(mod2)
```

8) Include both AB and ab as predictors and do a backward elimination
```{r}
mod3 <- lme(y ~ ab * AB, data = LME, random =  ~ 1 | id)
summary(mod3)
anova(mod3)# default  - type="s" sequential - tested in order they appear in model
anova(mod3, type = "m")#type = "m" marginal  - delete one term from the model at a time.
#interaction not significant
mod4 <- lme(y ~ ab + AB, data = LME, random =  ~ 1 | id)
summary(mod4)
anova(mod4, type = "m") #AB not significant
mod2 <- lme(y ~ ab, data = LME, random =  ~ 1 | id)
anova(mod2)

ggplot(LME, aes(x = interaction(ab, AB), y = y)) + 
  geom_boxplot()

```


9) What type of contrast is used?
```{r}
contrasts(LME$AB)#first column of 1s omitted
```

10) What is the expected mean for the different levels of AB and ab?
```{r}
summary(mod4)
coef(mod4)
fixef(mod4) #extract fixed effects
fixef(mod4)[1] #aA
fixef(mod4)[1] + fixef(mod4)[2] #bA
fixef(mod4)[1] + fixef(mod4)[3] #aB
fixef(mod4)[1] + fixef(mod4)[2] + fixef(mod4)[3] #bB
tapply(LME$y, list(LME$ab, LME$AB), mean)#different because no interaction fitted

library("dplyr")
LME %>%
  group_by(ab, AB) %>%
  summarise(mean = mean(y))
```

11) Use a linear model with AB and compare the results to question 4
```{r}
mod5 <- lm(y ~ AB, data = LME)
summary(mod5)
anova(mod5)
```

12) Use a linear model with ab and compare the results to question 6
```{r}
mod6 <- lm(y ~ ab, data = LME)
summary(mod6)
```


15) Import data length.csv <br>
id is the individual id, day is the predictor, length is the response
```{r}
if(interactive()){
  path <- ""
}else{
  path <- "../../"
}
leng <- read.csv(paste0(path, "data/length.csv"))
summary(leng)
```


16) Find the best model to explain the data
```{r}
mod7 <- lme(len ~ days, data = leng, random =  ~ days | id)
mod8 <- lme(len ~ days, data = leng, random =  ~ 1 | id)
anova(mod7, mod8)
anova(mod8)
```

```{r}
leng2 <-cbind(leng,
  pred1 = predict(mod8, level = 1),
  pred0 = predict(mod8, level = 0)
  )

ggplot(leng2, aes(x = days, y = len, group = id)) +
  geom_line() +
  geom_line(aes(y = pred1), colour = "blue") +
  geom_line(aes(y = pred0), colour = "red", size = 2)
```