---
title: "Practical 3"
author: "Richard J. Telford"
date: "May 27, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Does heteroscedastic matter

1. simulate some data with no relationship between the predictor and the response, but with changing variance
 - hint use `x <- 1:100`
 - `y <- rnorm(100, mean = 0, sd = seq(1, 10, length = 100))`
 
2. Fit a model. Is it significant?
```{r}
n <- 100
x <- 1:n
y <- rnorm(n, mean = 0, sd = seq(1, 10, length = n))

mod <- lm(y ~x)
library("ggfortify")
autoplot(mod)
anova(mod)
names(anova(mod))
anova(mod)$Pr[1]
```



3. Repeat this 1000 times. How many models are significant? How many would you expect?
 - hint use `replicate` extract p value from `anova`
```{r}
res <- replicate(1000, {
  y <- rnorm(n, mean = 0, sd = seq(1, 10, length = n))
  mod <- lm(y ~x)
  anova(mod)$Pr[1]
})
ggplot(data.frame(res), aes(x = res)) + 
  geom_histogram(binwidth = 0.05, center = 0.025) + 
  geom_hline(yintercept = 50)
sum(res < 0.05)
```


# NLS

- Import data amount.csv
```{r}
if(interactive()){
  amount <- read.csv("data/amount.csv")
}else{
  amount <- read.csv("../../data/amount.csv")
}
knitr::kable(head(amount))
```


- Do a non-linear regression

```{r}
library(nlme)
fit.nls<-nls(amount ~ b0 + b1 * exp(b2 * calcium), data = amount, 
             start = c(b0 = 0, b1 = 20, b2 = -1))
summary(fit.nls)

```

- Interpret the results
- What is the expected value if calcium = 10?

```{r}
coef(fit.nls)["b0"] + coef(fit.nls)["b1"] * exp(coef(fit.nls)["b2"] * 10)
predict(fit.nls, newdata = data.frame(calcium = 10))

ggplot(cbind(amount, fit = fitted(fit.nls)), aes(x = calcium, y = amount)) +
  geom_point() +
  geom_line(aes(y = fit))

```

