---
title: "Maximum likelihood"
author: "Richard J. Telford"
date: "May 27, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Maximum Likelihood

1.Import bird1.csv
```{r}
if(interactive()){
  bird1 <- read.csv("data/bird1.csv")
}else{
  bird1 <- read.csv("../../data/bird1.csv")
}
```

2.Fit a model using lm
```{r}
mod1 <- lm(weight ~ temp, data = bird1)
summary(mod1)
```
3.Pick a point and calculate the likelihood 
```{r}
s2 <- sum(resid(mod1) ^ 2) / mod1$df.residual
s2

L1 <- 1 / sqrt(2 * pi * s2)
L2 <- exp((-1 / (2 * s2)) * (bird1$weight[13] - fitted(mod1)[13]) ^ 2)
L1 * L2

dnorm(bird1$weight[13], mean = fitted(mod1)[13], sd = sqrt(s2))

plot(seq(0, 40, length = 100),
     dnorm(seq(0, 40, length = 100), mean = fitted(mod1)[13], sd = sqrt(s2)),
     type = "l")
abline(v =  bird1$weight[13])
abline(h = dnorm(bird1$weight[13], mean = fitted(mod1)[13], sd = sqrt(s2)))
```

4.For the same point estimate the log-likelihood
```{r}
log(dnorm(bird1$weight[13], mean = fitted(mod1)[13], sd = sqrt(s2)))
```

5.Estimate the log-likelihood for all points
```{r}
L2 <- exp((-1 / (2 * s2)) * (bird1$weight - fitted(mod1)) ^ 2)
sum(log(L1 * L2))

(lik <- sum(log(
  dnorm(bird1$weight, mean = fitted(mod1), sd = sqrt(s2))
  )))
```

6.Use different coefficients and redo question 2 to 4
```{r}
b0 <- 12
b1 <- 0.5
new.fit <- b0 + b1 * bird1$temp
new.resid <- bird1$weight - new.fit
new.s2 <- sum(new.resid ^ 2) / mod1$df.residual

(new.lik <-sum(log(
  dnorm(bird1$weight, mean = new.fit, sd = sqrt(new.s2))
)))
```


7.Use the lm and glm to estimate the relationship of weight and temperature

```{r}
mod2 <- glm(weight ~ temp, data = bird1)
```

8.Compare the results

```{r}
summary(mod1)
summary(mod2)
```

9.Calculate the likelihood for the saturated model (mu = y)

```{r}
(sat <- sum(log(
  dnorm(bird1$weight, mean = bird1$weight, sd = sqrt(s2))
  )))
```

10.Use the log-likelihoods to estimate the Deviance

```{r}
(D <- -2 * (lik - sat))
```

11.What is the relationship between the Deviance and the sum of squared residuals, assuming Normal distribution?
```{r}
sum(resid(mod1) ^ 2)
```
scaled deviance
```{r}
D * s2
```