---
title: "Linear models"
author: "Richard J. Telford"
date: "May 26, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Linear Modelling Exercises

1. Import the data found in bird1.csv
```{r}
if(interactive()){
  bird1 <- read.csv("data/bird1.csv")
}else{
  bird1 <- read.csv("../../data/bird1.csv")
}knitr::kable(head(bird1))
```

Always use relative paths. Absolute paths are machine dependent: not reproducible.

2. Fit a linear model, and interpret the coefficients.
```{r}
mod <- lm(weight ~ temp, data = bird1)
summary(mod)
```
Positive intercept - mean weight at zero temp is positive
Positive slope - heavier at higher temperatures

3. Examine the diagnostics plot. Is everything OK?
```{r}
par(mfrow = c(2, 2))
plot(mod)
```

With ggplot - first install extra packages if not present

```{r}
if(!require("devtools")) install.packages("devtools")
if(!require("ggfortify")) devtools::install_github("sinhrks/ggfortify")
#ggfortify is not currently on CRAN
#more info on ggfortify at http://rpubs.com/sinhrks/plot_lm
```

```{r}
autoplot(mod, label.size = 3)
```

3. Calculate the residual sum of squares (find sum the squared difference between the estimates calculated from the coefficients and the observed values).
```{r}
#find value with anova()
anova(mod)
#now calculate
modFit <- coef(mod)[1] + coef(mod)[2] * bird1$temp
modResid <- bird1$weight - modFit
sum(modResid^2)
```

4. Recalculate the Residual sum of squares using different values for the coefficients (in the formula yi = Beta0 + Beta1 * xi + εi).
```{r}
newFit <- 17 + -1 * bird1$temp
newResid <- bird1$weight - newFit
sum(newResid^2)
```

5. Is the new RSS larger or smaller?

Much larger!

6. Estimate the slope using covariance and variance.
```{r}
(beta1 <- with(bird1, cov(weight, temp)/sqrt(var(temp) * var(temp))))
(beta0 <- with(bird1, mean(weight) - beta1 * mean(temp)))
```

7. Do you get the same result?
```{r}
all.equal(as.vector(coef(mod)[1]), beta0)
all.equal(as.vector(coef(mod)[2]), beta1)
```

8. Create a linear model (one-way anova) with the factor of C & T as the predictor
```{r}
#plot graph first
ggplot(bird1, aes(x = treat, y = weight)) + geom_boxplot()
mod2 <- lm(weight ~ treat, data = bird1)
```



9. Compare the SSTot with the regression model.
```{r}
sum(anova(mod)$"Sum Sq")
sum(anova(mod2)$"Sum Sq")

```

10. Are there any difference.

Obviously not. SSTot is property of the data, not of the model.

12. How should the coefficients be interpreted?

 - First coefficient is mean of first factor level
 - Second coefficient is difference between means of first and second level

13. Import the data found in bird2.csv
```{r}
bird2 <- read.csv("../../data/bird2.csv")
knitr::kable(head(bird2))
```

14. Fit a linear model with both temperature and gender as predictors.
```{r}
mod3 <- lm(weight ~ temp * gender, data = bird2)
summary(mod3)
```

15. How should we interpret the results?

 - First coefficient - female intercept 
 - Second coefficient - effect of temp on females
 - Third coefficient - difference of male intercept from female intercept
 - Fourth coefficient - difference of effect on males from effect on females

16. What is the difference in intercept between male and female birds?
```{r}
coef(mod3)[3]
```


17. How much does the slope differ between male and female?
```{r}
coef(mod3)[4]
```

18. Is this significant?

Yes

```{r}
with(bird2, {
  plot(temp, weight, col = c(2, 4)[gender])
  lapply(levels(gender), function(g)
    lines(temp[gender == g], fitted(mod3)[gender == g], col = ifelse(g == "female", 2, 4)))
  legend("bottomright",legend = c("female", "male"), col = c(2, 4), lty = 1, pch = 1)
})

ggplot(cbind(bird2, fit = fitted(mod3)), aes(x = temp, y = weight, colour = gender)) +
  geom_point() +
  geom_line(aes(y = fit))
```
