---
title: "Bio302 Practical 1 Exploratory Data Analysis"
author: "Richard J. Telford"
date: "May 25, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath('../../'))
knitr::opts_chunk$set(echo = TRUE)
```


Download file combinedETE.csv from the git and place in a subdirectory called data of your project


Import the data into R (hint use read.csv(…))

```{r}
#Code assumes that rmd file is in "practical/solutions/" and data are in "data/"
dat <- read.csv("data/combinedETE.csv")
```

Find the range, mean, median and variance of variable percAgg.

```{r}
#range
range(dat$percAgg)

#mean
(paMean <- mean(dat$percAgg))

#median
(paMedian <- median(dat$percAgg))

#variance
var(dat$percAgg)

```

What is the standard error of the mean of percAgg.
```{r}
sd(dat$percAgg) / sqrt(length(dat$percAgg))
```

Plot a histogram of variable percAgg.
Use abline to mark the position of the mean and median.
```{r}
hist(dat$percAgg)
abline(v = c(paMean, paMedian), col = c(2, 4))
```
With `ggplot2`

```{r}
library("ggplot2")
g <- ggplot(dat, aes(x = percAgg)) + 
         geom_histogram() +
         geom_vline(xintercept = c(mean = paMean, median = paMedian), colour = c(2, 4))
print(g)       
```


Plot percAgg against numSites
```{r}
with(dat, plot(numSites, percAgg))
```


Find the covariance and correlation of percAgg and numSites
```{r}
with(dat, cov(percAgg, numSites))
with(dat, cor(percAgg, numSites))
```

Is the correlation significant? (hint use cor.test())

```{r}
with(dat, cor.test(percAgg, numSites))
```

How should the p-value be interpreted?

There are three levels in factor land.island. Plot percAgg against numSites coloured by the value of land.island.

```{r}
with(dat, plot(numSites, percAgg, col = land.island))

```
with ggplot
```{r}
g <- ggplot(dat, aes(x = numSites, y = percAgg, colour = land.island)) +
  geom_point() +
  geom_smooth(aes(group = 1), show.legend = FALSE)
print(g)
```


Is the correlation between percAgg and numSites significant for any of the groups?
```{r}
with(dat[dat$land.island == "ETE", ], cor(percAgg, numSites))
with(subset(dat, land.island == "ETE"), cor(percAgg, numSites))
```


```{r}
by(dat, dat$land.island, function(x)cor.test(x$percAgg, x$numSites))
```


This is a sub-group analysis, what are the dangers of this type of analysis?